---
title: "Historical Emissions (1990-2019) in R Marckdown"
author: "Jorge Escobar _ jorge.fsa23@gmail.com"
date: "First created on: Dec 17, 2022. Last compiled: `r format(Sys.time(), '%b %d, %Y %H:%M:%S')`"
output:
  html_document:
      highlight: zenburn
      theme: readable
      toc: yes
      toc_float:
         collapsed: yes
         smooth_scroll: yes 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- Personalizing CSS style -->
```{css, echo = FALSE}

p{
  text-align: justify;
  color: #696969
  }

h1{
  background-color: #f2e392;
  text-align: center;
  color: #5d0e16
}

h2,h3,h4,h5{
  background-color: #f1f5ad;
  text-align: left;
  color: #0000ff
}

```
## Introduction

This is an `R Marckdown` report developed in `R` language and `RStudio IDE`.

Database: 'historical_emissions.csv'

The Report was developed by data available in  [GHG_Emissions](https://www.climatewatchdata.org/ghg-emissions) behind the story *CO2 emissions (metric tons per capita)*.

Source: Climate Watch Historical GHG Emissions. 2022.
Washington, DC: World Resources Institute.

According to [ADEC-SOLUTIONS](https://www.adecesg.com/resources/faq/what-is-ghg/#:~:text=Greenhouse%20gases%2C%20or%20GHGs%2C%20are,these%20gases%20and%20the%20atmosphere):

Greenhouse gases (GHGs), are compound gases that trap heat or longwave radiation in the atmosphere, making the Earth’s surface warmer. 

The GHGs molecular structure allows them to absorb the heat released or trap them in the atmosphere and re-emit them back to the earth. 
This heat-trapping phenomenon is known as the greenhouse effect. 
The accumulation of GHGs since the industrial revolution has accelerated this greenhouse effect, causing global warming and climate change.

The principal GHGs, also known as heat-trapping gases, are carbon dioxide, methane, nitrous oxide, and the fluorinated gases. 
Carbon dioxide composes 64.3% of GHGs. 
It enters the atmosphere through the burning of fossil fuels, solid waste, trees and wood products, and certain chemical reactions. 
It is removed from the atmosphere naturally by plants as part of the biological carbon cycle.

lucf :: Land-use Change and Forestry


Glossary (database):
Country: Name of country/territory.
Sector: Total including LUCF (Land-use Change and Forestry).
Gas: All GHG (Including all kind of Greenhouse gases).
Unit: MtCO2e (Metric tons of carbon dioxide equivalent)
Data source: CAIT (Climate Watch)
Years: 1990-2019

>On the next sections, the results of basic data analysis with R.
	
## Preparing environment
 
### Loading packages
<!-- If necessary, the packages must be installed before in RStudio -->
```{r message=FALSE, warning=FALSE}
library(tidyverse) # data treatment
```

### Reading Database
```{r}
historical_emissions <- read_csv(file = "historical_emissions.csv")
```

## Structure

Function 'glimpse()' of the package dplyr (part of tidyverse)
```{r}
glimpse(historical_emissions)
```
We can observe that the first columns (Country, Data source, Sector, Gas, and Unit) where properly identified
as *character* and most of the years values as *double* a kind of numeric value. There is an exception with
the year 1990, where data was identified as *character*.

We must correct this whit the function 'mutate'
```{r}
historical_emissions <- historical_emissions %>% 
                            mutate_at(c("1990"), as.numeric)
  head(3) 
```
To verify last change
```{r}
glimpse(historical_emissions)
```
### Visualize dimension (column x rows)
Funtion 'dim()'
```{r}
dim(historical_emissions)
```
That is, We have 195 countries/regions on this database
Function 'nrow()' shows number of rows
```{r}
nrow(historical_emissions)
```
Function 'ncol()' shows number of columns
```{r}
ncol(historical_emissions)
```
Visualize names of variables into database (columns)
Function 'colnames()'
```{r}
colnames(historical_emissions)
```
Visualize data base. Function 'View()'
```{r}
View(historical_emissions)
```
We observed that some columns do not have relevant information
We can retire those variables (Sector', 'Data source', 'Gas', 'Unit')
Function 'select()' of the package dplyr
```{r}
historical_emissions <-  historical_emissions %>% 
                            select(-'Sector', - 'Data source', -'Gas', -'Unit')
```

Visualize database to confirm
```{r}
glimpse(historical_emissions)
```

To organize data by most pollutant countries
Function 'arrange()' of package: dplyr
Lets see the most pollutant countries in 1990
```{r}
arranged1990 <- historical_emissions %>% 
                        arrange(desc(historical_emissions$`1990`))
```
To observe the top 10 countries in 1990:
As we has "World" as compiling all countries, we need to print the top 11 on the list
```{r}
arranged1990 %>%
  select(c(Country, `1990`)) %>%
  print(n=11)
```

Lets see the Countries organized originally (by 2019 values)
```{r}
arranged2019 <- historical_emissions %>% 
                  arrange(desc(historical_emissions$`2019`))
```

To observe the top 10 countries in 2019:
As we has "World" as compiling all countries, we need to print the top 11 on the list
```{r}
arranged2019 %>%
  select(c(Country, `2019`)) %>%
  print(n=11)
```
Comparing both tables we observe differences between countries...


To identify emission > 1000 in 2019 (filtering by x values...)
```{r}
filtered_emissions <- historical_emissions %>% 
              filter(historical_emissions$`2019` > 1000.0) %>% 
              select(c(Country, `2019`)) %>%
              print(n=21)
```

To identify emission > 1000 in 1990
```{r}
filtered_emissions <- historical_emissions %>% 
                          filter(historical_emissions$`1990` > 1000.0) %>% 
                          select(c(Country, `1990`)) %>%
                          print(n=21)
```

To identify a specific country (ex. Colombia)
Function 'filter()' of package: dplyr
```{r}
historical_emissions %>% 
  filter(Country == "Colombia") %>%
  select(Country, `2019`, `2018`)
```

We can create a simple classification based on levels.
Ex. we can choose year 2019 and insert a new column with levels A, B, C, D
Being D to the higher and A to the lower.
Funcao 'mutate()' do pacote dplyr
```{r}
emissions_levels2019 <- historical_emissions %>% 
                          mutate(class_levels = 
                                   ifelse(`2019` < 100, "A", 
                                          ifelse(`2019` >= 100 & `2019` < 500, "B",
                                                 ifelse(`2019` >= 500 & `2019` < 1000, "C",
                                                        "D")))) %>%
                          select(Country, `2019`, class_levels)
```
Visualising this new change...
```{r}
view(emissions_levels2019)
```

Now, its is possible to visualize the number of countries on each level
```{r}
emissions_levels2019 %>% 
    count(class_levels)
```
So, most of the countries are in lower (A), 42 in level B, 9 in C and 9 in D

To visualize the mean of emission for a year 
Comparing 2019 and 1990).
Function 'summarise()' of package dplyr
```{r}
historical_emissions %>% 
  summarise(mean_2019 = mean(`2019`))
```
```{r}
historical_emissions %>% 
  summarise(mean_2019 = mean(`1990`, na.rm = TRUE))
```
In 1990 the mean was 351, while this value increased to 517.


To show the number of countries (values) considered to calculate the mean 
```{r}
historical_emissions %>% 
    summarise(mean_2019 = mean(`2019`, na.rm = TRUE), n = n())
```
The value of 195 match with the number of rows (Countries) into the table

To see the mean to more years..., ex. last 8 years (and arrange into a new db)
```{r}
historical_mean <- historical_emissions %>% 
                      summarise(m_2019 = mean(`2019`),
                                m_2018 = mean(`2018`),
                                m_2017 = mean(`2017`),
                                m_2016 = mean(`2016`),
                                m_2015 = mean(`2015`),
                                m_2014 = mean(`2014`),
                                m_2013 = mean(`2013`),
                                m_2012 = mean(`2012`))
```

# To visualize the new database
```{r}
View(historical_mean)
```

To observe more estatistical data for a year (2019).
Mean, Standard dev., sum, median, variance, max, min and length values
```{r}
historical_emissions %>% 
  summarise(m_2019 = mean(`2019`),
            sd_2019 = sd(`2019`),
            sum_2019 = sum(`2019`),
            median = median(`2019`),
            variance = var(`2019`),
            max_value = max(`2019`),
            min_value = min(`2019`),
            length_2019 = length(`2019`))
```
## Conclusion 
